---
title: "Estimation with Right-Censored Event-Time Data"
author: "Dave Harrington"
date: "May 14 - 18, 2018"
output: 
  beamer_presentation:
     includes:
       in_header: ../survival_header.tex
     fig_width: 3.25
     fig_height: 3.0
     fig_caption: true
     toc: true
     keep_tex: false
slide_level: 3
urlcolor: darkblue 
linkcolor: darkblue
citecolor: darkblue
--- 

# The Kaplan-Meier estimator

### Approaches to estimating $S(t)$

- Parametric models and maximum likelihood

- *The non-parametric Kaplan-Meier (KM) estimate*
    - KM also called the product limit estimator because of original derivation


### The Kaplan-Meier Estimator: General idea

The Kaplan-Meier estimator is probably the most
popular approach. 

When there is no censoring, the general formula is:
\[
\widehat{S}(t) = \frac{\#~individuals~ with~T >  t}{total~sample~size}
\]

### An example: Cox and Oakes, no censoring

*Time to relapse* (weeks) for 21 leukemia patients
receiving control treatment\footnote{Table 1.1 of Cox \& Oakes, 1984}: 

- 1, 1, 2, 2, 3, 4, 4, 5, 5, 8, 8, 8, 8, 11, 11, 12, 12, 15, 17, 22, 23

What is $\widehat{S}(10) = \widehat{P}(T > 10)$, the probability that an individual survives more than 10 weeks?

- This is 8/21 = 0.38 since 8 people survive more than 10 weeks.

What about $\widehat{S}(8)$?

- $\widehat{S}(8) =  \widehat P(T > 8) = 8/21 = 0.38$
- The four events at $t = 8$ are counted as having already failed.


### Empirical survival function

When there is no censoring, the general formula is:
\begin{align*}
\widehat{S}(t) = \frac{\#~individuals~ with~T > t}
       {total~sample~size}
\end{align*}

What is the standard error of $\widehat{S}(t)$?

- When there is no censoring, the estimated survival function is a proportion $\hat{p}$ with the standard error:
\[
\text{s.e.}[\widehat{S}(t)] = \sqrt{p(1-p)/n}
\]

\[\text{Example: s.e.}[\widehat{S}(8)]= \sqrt{(0.38)(0.62)/21} = 0.106 \]

### A table of $\widehat{S}(t)$

\begin{center}
\begin{tabular}{c|c|c}
Values of t & \# individuals with $T > t$ & ~~~~$\widehat{S}(t)$~~~~ \\
\hline
$0 \leq t < 1 $  & 21 & 21/21=1.000 \\
$1 \leq t < 2 $  & 19 & 19/21=0.905\\
$2 \leq t < 3 $  & 17 & 17/21=0.809\\
$3 \leq t < 4 $  & 16 & 16/21=0.762\\
$4 \leq t < 5 $  & 14 & 14/21=0.667\\
$5 \leq t < 8 $  & 12 & 12/21=0.571\\
$8 \leq t < 11 $ & 8 & 8/21=0.381\\	
$11 \leq t < 12 $ & 6 & 6/21=0.286\\
$12 \leq t < 15 $ & 5 & 4/21=0.191\\
$15 \leq t < 17 $ & 3 & 3/21=0.143\\
$17 \leq t < 22 $ & 2 & 2/21=0.095\\
$22 \leq t < 23 $ & 1 & 1/21=0.048\\
\end{tabular}
\end{center}

<!---

\footnotesize
\begin{tabular}{r|lllllllllllllllllllll}
$T_i$ & 1& 1& 2& 2& 3& 4& 4& 5& 5& 8& 8& 8& 8& 11& 11& 12& 12& 15&
17& 22& 23\\
\hline
Order     & 1& 2& 3& 4& 5& 6& 7& 8& 9& 10& 11& 12& 13& 14& 15& 16& 17&
18& 19& 20& 21\\
\end{tabular}

--->

### What about censoring?

Consider time to relapse (weeks) for leukemia patients in the treatment group.\footnote{Table 1.1 of Cox and Oakes} Times with $^+$ are right censored:
\[6^+,6,6,6,7,9^+,10^+,10,11^+,13,16,17^+\]
\[19^+,20^+,22,23,25^+,32^+,32^+,34^+,35^+\]

Naturally, $\widehat{S}(6-)= 21/21$    

- because everyone survived until at least time 6 or greater   

Not right to claim $\widehat{S}(6) = 17/21$   

- due to unknown status of person censored at time 6


### Censoring with the Kaplan-Meier

In a 1958 paper in the *Journal of the American Statistical Association*, Kaplan and Meier proposed a way to nonparametrically estimate $S(t)$,
in the presence of censoring.  

The method is based on the ideas of *conditional probability*.

### Censoring and the KM estimator

$S(t)$ in the discrete case:

To estimate $S(t)$ for time $t$ within the interval $t_k$ and $t_{k+1}$, e.g. $t_k \leq t < t_{k+1}$, consider the intervals defined by the ordered $k$ failure times,
\[ 
[t_{0}, t_{1}), [t_{1}, t_{2}), \ldots, [t_{k-1}, t_{k}), [t_{k}, \infty)
\]

The KM estimate is constructed based on events within each interval $[t_{j}, t_{j+1})$

- $d_j$ is the number of deaths in the interval $[t_{j}, t_{j+1})$

- $r_j$ is the number of individuals at risk in the interval $[t_{j}, t_{j+1})$

Initial assumptions: $t_0=0, \;\; P(T > t_0) = 1.$

### Censoring and the KM: discrete case

Then,

\begin{align*}
S(t)=  P(T >t) &=  P(T > t_{k}) \\
   &=  P(T >  t_1, T > t_2,\ldots,T >  t_{k})\\
   &=  P(T >  t_1) \times \prod_{j=2}^k \, P(T >  t_{j}|T >  t_{j-1}) \\
&\stackrel{(*)}{=}  \prod_{j=1}^k \, [1-P(T=t_j|T >  t_{j-1})] = \prod_{j=1}^k \, [1-\lambda_j]\\
\mbox{so}~~~ \widehat{S}(t) &\cong  \prod_{j=1}^k \, \left(1-\frac{d_j}{r_j}\right) =
\prod_{j:t_j \leq t}\left(1-\frac{d_j}{r_j}\right)
\end{align*}

\footnotesize
(*) Initial assumptions: $t_0=0, \;\; P(T > t_0)=1.$ 

### Censoring and the KM: continuous case 

For continuous data, the Kaplan-Meier estimator of the survivorship function $S(t)=P(T > t)$ is
\[
\widehat{S}(t) = \prod_{j: \tau_j \leq t} \frac{ r_j - d_j } {r_j}
= \prod_{j: \tau_j \leq t} \left(1 - \frac{d_j}{r_j}\right), \text{where}
\]

- $\tau_1, \ldots, \tau_K$ are the K distinct death times observed
- $d_j$ is the number of deaths at $\tau_j$
- $r_j$ is the number of individuals ``at risk'' right before the $j$-th
death time (everyone dead or censored *at or after* that time).
    - $r_j = r_{j-1} - d_{j-1} - c_{j-1}$
    - Alternatively, $r_j = \sum_{l \ge j}(c_l+d_l)$
- $c_j$ is the number of censored observations between the $j$-th and $(j+1)$-th death times.
    - Censorings tied at $\tau_j$ are included in $c_j$



### Computing

Most widely used software packages (SAS, Stata, R) have modules for survival analysis.

We will focus on R since it is free and has very good survival routines written by Terry Therneau.


### Fitting a Kaplan-Meier in R

\scriptsize

```{r, echo=TRUE, eval=TRUE, fig.height = 4.2, fig.width = 7}
library(survival)
library(eventtimedata)
data("cox.oakes.leukemia")
leukemia.remission <- survfit(Surv(time, relapse) ~ group, 
                              data = cox.oakes.leukemia)
plot(leukemia.remission, lty = 2:3, mark.time = TRUE, xlab = "Weeks", 
     ylab = "Survival Function" )
```

### Numerical output

\footnotesize

```{r, echo=TRUE, eval=TRUE, tidy.opts=list(width.cutoff=40)}
library(survival)
library(eventtimedata)
print(leukemia.remission)
```


### KM numerical estimates, group == 0

\scriptsize

```{r echo = TRUE, eval=TRUE}
leukemia.group.0 = subset.data.frame(cox.oakes.leukemia, group == 0)
km.group.0 = survfit(Surv(time, relapse) ~ 1, data = leukemia.group.0)

summary(km.group.0)
```

### KM numerical estimates, group == 1

\scriptsize

```{r echo = TRUE, eval=TRUE}
leukemia.group.1 = subset.data.frame(cox.oakes.leukemia, group == 1)
km.group.1 = survfit(Surv(time, relapse) ~ 1, data = leukemia.group.1)

summary(km.group.1)
```
\normalsize

\vspace{1cm}

Subsets used here only to fit output on slides.

\texttt{summary(leukemia.remission)} prints values for both groups.

# Estimating standard errors

### Pointwise confidence intervals for the KM

Why *pointwise*? 

- Since the KM is a function of time, there is an estimate of the standard error (or the variance) at each time. 

*Greenwood's formula* is the most commonly used estimate of the KM standard error.
\[
\widehat{\text{var}} (\widehat{S}(t)) =
[\widehat{S}(t)]^2 \, \sum_{j: \tau_j \leq t} \frac{d_j}{(r_j-d_j) r_j}
\]

Derivation given later in the slides.


### Confidence intervals for the KM    

A 95\% confidence interval  could be based on
\[  
\widehat{S}(t)  \pm z_{1-\alpha/2} \times \text{s.e.}[\widehat{S}(t)], 
\]
with $\text{s.e.}[\widehat{S}(t)]$  estimated using Greenwood's formula.

- However, this approach can yield values $>1$ or $<0$.

The better approach is to use the *log-log* transformation and base intervals around 
\[
L(t) = \log[-\log[S(t)]]
\]

\vspace{1cm}

\footnotesize
In \textsf{R}, use the option \texttt{conf.type = "log-log"}. The default transformation in \textsf{R} is $L(t) = -\log[S(t)]$.


### Confidence intervals \ldots  

To transform back, use $S(t)=\exp[-\exp[L(t)]]$.

\vspace{1cm}

Since...

- $0 \leq S(t) \le 1$, 

-  $0 \leq -\log[S(t)] < \infty$, and

-  $-\infty < \log[-\log[S(t)]] < \infty,$

the confidence interval will be in the proper range when transformed back.

### Log-log approach for confidence intervals:

1. Define $L(t) = \log[-\log[S(t)]]$.

2. Form a 95\% confidence interval for $L(t)$, $(\widehat{L}(t)-A,\widehat{L}(t)+A),$   with $A= 1.96 \times \text{s.e.}[\widehat{L}(t)].$ 

3. Apply $S(t)=\exp[-\exp[L(t)]]$ to obtain the confidence bounds for the 95\% CI on $S(t)$, 
\[
\left( \exp[-e^{(\widehat{L}(t)+A)}],\exp[-e^{(\widehat{L}(t)-A)}]\right)
\]

4. Substituting $\widehat{L}(t)=\log[-\log[\widehat{S}(t)]]$ back into the above bounds yields confidence bounds of
\[
\left([\widehat{S}(t)]^{e^A},[\widehat{S}(t)]^{e^{-A}}\right)
\]

### Confidence intervals for median survival

The median is usually defined as 

\[
    q_{0.5} = \min\{t_j: \widehat{S}(t) \ge 0.5\}.
\]

Other quantiles are defined similarly.

Confidence limits for median survival are based on confidence intervals for $S(t)$.

R uses the method  due to Brookmeyer and Crowley (Biometrics 1982, 38, 29–41).    
- SAS and other packages use this as well.

The formulas are complex and not shown here.

# The cumulative hazard estimator

### Estimating *S(t)* via the Nelson-Aalen cumulative hazard

The cumulative hazard $\Lambda(t)$ can be approximated by a sum over $j$ intervals,
\[\Lambda(t) \approx \sum_{j}  \lambda_j  \Delta \]

where 

-  $\lambda_j$ is the value of the hazard in the $j^{th}$ interval 

-  $\Delta$ is the width of each interval

Since $\widehat{\lambda}_j \Delta$ is approximately the probability of having an event in an interval $j$, conditional on having survived until the beginning of the interval, $\Lambda(t)$ can be approximated further as
\[\Lambda(t) \approx \sum_{j}  \lambda_j  \Delta \approx \sum_{j} \frac{d_j}{r_j}  \]

### Estimating *S(t)* via the Nelson-Aalen cumulative hazard \ldots

Thus, the *Nelson-Aalen estimator* can be written as
\[\widehat{\Lambda}_{\tiny NA}(t) =  \sum_{t_j \leq t} \frac{d_j}{r_j} \]

From $\widehat{\Lambda}_{\tiny NA}(t)$, an alternative to the KM estimator of $S(t)$ can be calculated:

\[  \widehat{S}_{\tiny FH}(t) = \exp[-\widehat{\Lambda}_{\tiny NA}(t)] \]

The *Fleming-Harrington estimator* is generally very close to $\widehat{S}_{\tiny KM}(t)$.



### Example: Time to recidivism, Rossi (1980)

Recidivism is the event of rearrest and reincarceration after release from prison.

A randomized study\footnote{\texttt{rossi} dataset in \texttt{eventtimedata} package.} with 52 weeks of follow-up after randomization collected information on the following variables:

- \texttt{fin}:  Financial support vs no financial support after release

- \texttt{week}:  Time in weeks to either re-arrest or censoring

- \texttt{arrest}: \texttt{1} = arrest during the follow-up,  \texttt{0} = no arrest


### Important detail

Always check the order of categories in a factor variable in R when labeling plots

```{r}
levels(rossi$fin)
```

### KM of recidivism, with confidence intervals

\footnotesize

```{r, eval = FALSE, echo=TRUE, fig.width=4.5, fig.height=3.8}
library(survival)
library(eventtimedata)
data("rossi")
rossi.recidivism.km <- survfit(Surv(week, arrest) ~ fin, 
                               data = rossi)
plot(rossi.recidivism.km, lty = 2:3, col = 3:4, mark.time = TRUE, 
     xlab = "Weeks", 
     ylab = "Probability of No Recidivism", 
     axes = FALSE, 
     conf.times = c(10,30,50),
     main = "KM of No Recidivism Probability, with Conf. Int.",
     cex = 0.6, cex.main = 0.8)
axis(1)
axis(2)
legend(10, .5, c("No Financial Support", "Financial Support"), 
       lty = 2:3, col = 3:4, cex = 0.6)
```

----- 

```{r, eval = TRUE, echo=FALSE, fig.width=4.5, fig.height=3.8}
library(survival)
library(eventtimedata)
data("rossi")
rossi.recidivism.km <- survfit(Surv(week, arrest) ~ fin, 
                               data = rossi)
plot(rossi.recidivism.km, lty = 2:3, col = 3:4, mark.time = TRUE, 
     xlab = "Weeks", 
     ylab = "Probability of No Recidivism", 
     axes = FALSE, 
     conf.times = c(10,30,50),
     main = "KM of Recidivism Probability, with Conf. Int.",
     cex = 0.6, cex.main = 0.8)
axis(1)
axis(2)
legend(10, .5, c("No Financial Support", "Financial Support"), 
       lty = 2:3, col = 3:4, cex = 0.6)
```

### Cumulative hazard (risk) of recidivism, w/CIs

\footnotesize

```{r, eval = FALSE, echo=TRUE, fig.width=4.5, fig.height=3.8}
library(survival)
library(eventtimedata)
data("rossi")
rossi.recidivism.ch <- survfit(Surv(week, arrest) ~ fin, 
                               data = rossi)
plot(rossi.recidivism.km, lty = 2:3, col = 3:4, mark.time = TRUE, 
     fun = "cumhaz",
     xlab = "Weeks", 
     ylab = "Cumulative Hazard of Recidivism", 
     axes = FALSE, 
     conf.times = c(10,30,50),
     main = "Cumulative Risk of Recidivism, 
     with Conf. Int.",
     cex = 0.6, cex.main = 0.8)
axis(1)
axis(2)
legend("topleft", inset = c(0.1, 0.3), 
       c("No Financial Support", "Financial Support"),
       lty = 2:3, col = 3:4, cex = 0.6)
```

------

```{r, eval = TRUE, echo=FALSE, fig.width=4.5, fig.height=3.8}
library(survival)
library(eventtimedata)
data("rossi")
rossi.recidivism.ch <- survfit(Surv(week, arrest) ~ fin, 
                               data = rossi)
plot(rossi.recidivism.km, lty = 2:3, col = 3:4, mark.time = TRUE, 
     fun = "cumhaz",
     xlab = "Weeks", 
     ylab = "Cumulative Hazard of Recidivism", 
     axes = FALSE, 
     conf.times = c(10,30,50),
     main = "Cumulative Risk of Recidivism, 
     with Conf. Int.",
     cex = 0.6, cex.main = 0.8)
axis(1)
axis(2)
legend("topleft", inset = c(0.1, 0.3), 
       c("No Financial Support", "Financial Support"),
       lty = 2:3, col = 3:4, cex = 0.6)
```

### Confidence intervals vs confidence bands

Examining many confidence intervals may cause the same problem as simultaneous hypothesis tests.

- Overall coverage probability for the curve is not right.   

Hall and Wellner (*Biometrika*, 1980) solved that problem by deriving confidence bands:

- 95\% bands have probability 0.95 of covering the entire survival curve.

- These bands will be wider than pointwise intervals.

- Formulas complex, not shown here.


### KM of recidivism, with confidence bands

\footnotesize

```{r, eval = FALSE, echo=TRUE, fig.width=4.5, fig.height=3.8}
library(survival)
library(eventtimedata)
data("rossi")
rossi.recidivism.km <- survfit(Surv(week, arrest) ~ fin, 
                               data = rossi)
plot(rossi.recidivism.km, lty = 2:3, col = 3:4, mark.time = TRUE, 
     xlab = "Weeks", 
     ylab = "Probability of No Recidivism", 
     axes = FALSE, 
     conf.int = TRUE,
     main = "KM of Probability of No Recidivism, with Conf. Bands",
     cex = 0.6, cex.main = 0.8)
axis(1)
axis(2)
legend(10, .5, c("No Financial Support", "Financial Support"),
       lty = 2:3, col = 3:4, cex = 0.6)
```

-----

```{r, eval = TRUE, echo=FALSE, fig.width=4.5, fig.height=3.8}
library(survival)
library(eventtimedata)
data("rossi")
rossi.recidivism.km <- survfit(Surv(week, arrest) ~ fin, 
                               data = rossi)
plot(rossi.recidivism.km, lty = 2:3, col = 3:4, mark.time = TRUE, 
     xlab = "Weeks", 
     ylab = "Probability of No Recidivism", 
     axes = FALSE, 
     conf.int = TRUE,
     main = "KM of Probability of No Recidivism, with Conf. Bands",
     cex = 0.6, cex.main = 0.8)
axis(1)
axis(2)
legend(10, .5, c("No Financial Support", "Financial Support"),
       lty = 2:3, col = 3:4, cex = 0.6)
```

### Example: Application to FDA (7 March 2018)


On 7 March 2018, Amgen asked for FDA approval of the drug blinatumumab in patients with a sub-type of acute lymphoblastic leukemia (ALL).

- The drug would be given to patients who experienced a clinical complete remission, but had evidence of minimal residual disease (MRD).

Figure on the next slide from the [FDA analysis](https://www.fda.gov/downloads/AdvisoryCommittees/CommitteesMeetingMaterials/Drugs/OncologicDrugsAdvisoryCommittee/UCM599298.pdf) of the data shows

- Relapse free survival by MRD status

- Shows confidence bands (Hall and Wellner)

----

![FDA presentation, 7 March 2018](../figures/usfda_mrd_leukemia_km.pdf){width=90%}

# Derivations

### KM estimator derivation, continuous case \ldots

*Conditional Probability:* Suppose $A$ and $B$ are two events. Then,
\[  P(A|B)  =    \frac { P(A \cap B)}  { P(B)}\]

*Multiplication Law*: Multiply both sides of the above by $P(B)$.
\[P(A \cap B) =  P(A|B) P(B) \]

*Extension to more than 2 events:* Suppose $A_1,  A_2, \ldots, A_k$ are $k$ different events. Then, the probability of all $k$ events occurring can be
written as a product of conditional probabilities.
\begin{align*}
P(A_1 \cap A_2 \ldots \cap A_k) =& \ P(A_k|A_{k-1} \cap \ldots \cap A_1) \\
& \qquad \times P(A_{k-1}|A_{k-2} \cap \ldots \cap A_1) \\
& \qquad \times \ldots \\
& \qquad \times P(A_2 | A_1) \\
& \qquad \times P(A_1)
\end{align*}

$$
   P(A_1 \cap A_2 ... \cap A_k)   = $$ 
   
$$ P(A_k|A_{k-1} \cap ... \cap A_1)
         \times  P(A_{k-1}|A_{k-2} \cap  ... \cap A_1) \times
              ...     \times P(A_2| A_1) \times P(A_1) .$$

### KM estimator derivation, continuous case \ldots

Think of dividing the observed time-span of the study into
a series of small intervals so that there is a separate
interval for each time of death or censoring (with possible ties):
\begin{center}
\begin{tabular}{l|p{.4in}|p{.4in}|p{.4in}
 |p{.4in}|p{.4in}|p{.4in}|p{.4in}|p{.4in}|p{.4in}|p{.4in}
 |p{.4in}|p{.4in}|p{.4in}}
  &   &  &   & &   & &   & C &   &   \\
  &   &  & D & & C & & C & D & D & D \\  \hline
\end{tabular}
\end{center}

Using the law of conditional probability,
\[P(T > t) = \prod_j P(\text{survive $j$-th interval $I_j$ | survived to start of $I_j$}), \]
over all intervals preceding time $t$.

### KM estimator derivation, continuous case \ldots

*Four possibilities for each interval:*

1. No event: conditional probability of surviving the interval is 1. 

2. Censoring: assume individual survives to end of the interval, so that the conditional probability of surviving the interval is 1.
     
3. Death, but no censoring: conditional probability of  *not* surviving the interval is \# deaths ($d$) divided by \# "at risk" ($r$) at the beginning of the interval. Thus, the conditional probability of surviving the interval is $1 - \frac{d}{r}$.
     
4. Tied deaths and censoring: assume censorings survive to end of the interval, so that conditional probability of surviving the interval is still $1 - \frac{d}{r}$.

Thus, the general formula for the conditional probability of surviving the $j$-th interval that holds for all 4 cases is $1-\frac{d_j}{r_j}$.
     

### KM estimator derivation, continuous case \ldots

As the intervals become smaller, 

- The approximations made in estimating the probabilities of surviving each interval become smaller.    
- The estimator converges to the true $S(t)$ as the sample size increases.

This argument clarifies why an alternative name for the KM is
the  *product limit estimator*.

### Result stated earlier

For continuous data, the Kaplan-Meier estimator of the survivorship function $S(t)=P(T > t)$ is
\[
\widehat{S}(t) = \prod_{j: \tau_j \leq t} \frac{ r_j - d_j } {r_j}
= \prod_{j: \tau_j \leq t} \left(1 - \frac{d_j}{r_j}\right), \text{where}
\]

- $\tau_1, \ldots, \tau_K$ are the K distinct death times observed
- $d_j$ is the number of deaths at $\tau_j$
- $r_j$ is the number of individuals ``at risk'' right before the $j$-th
death time (everyone dead or censored *at or after* that time).
    - $r_j = r_{j-1} - d_{j-1} - c_{j-1}$
    - Alternatively, $r_j = \sum_{l \ge j}(c_l+d_l)$
- $c_j$ is the number of censored observations between the $j$-th and $(j+1)$-th death times.
    - Censorings tied at $\tau_j$ are included in $c_j$

### Derivation of Greenwood's formula

KM estimator can be thought of as 
\[  
\widehat{S}(t) = \prod_{j: \tau_j \leq t} (1-\widehat{\lambda}_j), \text{ where } \widehat{\lambda}_j = \frac{d_j}{r_j}.
\]

Since the $\widehat\lambda_j$'s are (conditionally) binomial proportions, standard likelihood theory can be used to to show each $\widehat{\lambda}_j$ is approximately normally distributed, with mean $\lambda_j$, and variance\footnote{The estimated variance is $\widehat{\text{var}}(\widehat{\lambda}_j) = \frac{\widehat{\lambda}_j
(1-\widehat{\lambda}_j)}{r_j}$.}
\[
\text{var}(\widehat{\lambda}_j) 
= \frac{{\lambda}_j (1-{\lambda}_j)}{r_j}
\]

The $\widehat\lambda_j$'s are independent in large enough samples.

### Derivation of Greenwood's formula \ldots

Since $\widehat{S}(t)$ is a function of the $\lambda_j$'s, its variance can be estimated using the *delta method*,

- an approach for calculating the variance of non-linear functions.

*Delta method:* If $Y$ is normal with mean $\mu$ and variance $\sigma^2$,
then $g(Y)$ is approximately normally distributed with 
mean $g(\mu)$ and variance $[g'(\mu)]^2 \sigma^2$.


### Digression: the delta method

Two specific examples that will be used in the derivation:

- Ex. 1: $Z = g(Y)= \log(Y)$, then $g'(y)=(1/y)$:
\[
Z \sim N\left(\log(\mu),\left(\frac{1}{\mu}\right)^2 \sigma^2\right)
\]

- Ex. 2: $Z = g(Y)= \exp(Y)$, then $g'(y)=e^y$:
\[
Z \sim N\left(e^{\mu}, [e^{\mu}]^2 \sigma^2\right)
\]

### Derivation of Greenwood's formula \ldots

Instead of dealing with $\widehat{S}(t)$ directly, use $\log[\widehat{S}(t)]$ since calculating variance of a sum is easier than calculating variance of a product,
\[ \log[\widehat{S}(t)] = \sum_{j: \tau_j \le t} \log(1-\widehat{\lambda}_j)  \]

By approximate independence of the $\widehat{\lambda}_j$'s,
\[\text{var}(\log[\widehat{S}(t)]) = 
\sum_{j: \tau_j \le t} \text{var}[\log(1-\widehat{\lambda}_j)]. \]


\small

Apply the delta method (Ex. 1), where $\mu = 1 - \lambda_j$ and $\sigma^2 = \frac{{\lambda}_j (1-{\lambda}_j)}{r_j}$.

\vspace{-0.4cm}

\begin{align*}
\widehat{\text{var}}(\log[\widehat{S}(t)]) =& \sum_{j: \tau_j \leq t} \left(\frac{1}{1 - \widehat{\lambda}_j} \right)^2 \left(\frac{\widehat{{\lambda}}_j (1-\widehat{{\lambda}}_j)}{r_j} \right) \\
=& \sum_{j: \tau_j \leq t} \frac{\widehat{\lambda}_j}{(1-\widehat{\lambda}_j) r_j} = \sum_{j: \tau_j \leq t} \frac{d_j}{(r_j-d_j) r_j}
\end{align*}

### Greenwood's formula

To obtain $\widehat{\text{var}}(\widehat{S}(t))$, apply the delta method again (Ex. 2), using the relationship $\widehat{S}(t) = \exp[\log[\widehat{S}(t)]]$,
\[\widehat{\text{var}}(\widehat{S}(t)) = [\widehat{S}(t)]^2 \ \widehat{\text{var}}\left[\log[\widehat{S}(t)] \right] \]

Substitute the previous result for $\widehat{\text{var}}\left[\log[\widehat{S}(t)] \right]$ to obtain Greenwood's Formula,
\[\widehat{\text{var}}(\widehat{S}(t)) = [\widehat{S}(t)]^2  \sum_{j: \tau_j \leq t} \frac{d_j}{(r_j-d_j) r_j} \]



