---
title: "Introduction and Background"
author: "Dave Harrington"
date: "May 14 - 18, 2018"
output: 
  beamer_presentation:
     includes:
       in_header: ../survival_header.tex
     fig_width: 3.25
     fig_height: 3.0
     fig_caption: true
     toc: true
     keep_tex: true
slide_level: 3
urlcolor: darkblue 
linkcolor: darkblue
citecolor: darkblue
--- 

# Administrative Information

### Instructor coordinates

Dave Harrington 

- Department of Biostatistics, Harvard T.H. Chan School of  Public Health  
- Department of Biostatistics and Computational Biology, Dana-Farber Cancer Institute
- Email: \href{mailto:dharrington846@gmail.com}{dharrington846@gmail.com}



### Course format

- Lectures, but please ask questions
- Labs (practical exercises)
- Readings (typically, papers from literature)
- Computing (in \textsf{R})

### Acknowledgements

Many individuals have contributed to these notes in courses at Harvard, Ghent University, Hasselt University, and other institutions.  

Special acknowledgements to    

- Rui Wang, Paige Williams, Rebecca Betensky, and Paul Catalano (Harvard)    
- Louise Ryan (University of Technology, Sydney)
- Els Goetghebeur (U Ghent)   
- Julie Vu (University College London)

# Important Definitions

### Main ideas

Survival analysis typically focuses on *time-to-event* data.

Examples:

- Time to death from a chronic disease
- Time to progression of a disease
- Time to onset (or relapse) of a disease
- Length of stay in a hospital or a nursing home

The terms *survival time* and *event time* will be used to mean the time to an event. 

### Most useful references

- Klein and Moeschberger: *Survival Analysis: Techniques for censored and truncated data*
- Therneau and Grambsch: *Modeling Survival Data: extending the Cox Model (R)*
- Collett: *Modelling Survival Data in Medical Research*
- Hosmer, Lemeshow, and May: *Applied Survival Analysis, 2nd ed.*
- Kleinbaum: *Survival Analysis:  A self-learning text*
- Cox and Oakes: *Analysis of Survival Data*

I recommend the K\&M text, supplemented by T\&G.

### A few more references

- Fleming and Harrington: *Counting Processes and Survival Analysis*
- Kalbfleisch and Prentice: *The Statistical Analysis of Failure Time Data*
- Allison: *Survival Analysis Using the SAS System*
- Miller: *Survival Analysis*


### Packages used

\scriptsize

```{r, echo=TRUE, eval=FALSE}
install.packages("devtools")
library(devtools)
install.packages(c("survival", "KMSurv",
                   "gsDesign", "Hmisc"))
install_github("keaven/nphsim", 
               "dave-harrington/eventtimedata")
```

### Example: Time to death or hospitalization

![Figure from Marrouche, et al., *NEJM* 2018](../figures/atrial_fib_death_hosp.pdf){width=70%}

See [Marrouche, et al.](../../clinical_papers/catheter_ablation_atrial_fib.pdf)


### Structure of event time data

```{r, echo = FALSE, warning = FALSE, out.width = "50%", fig.cap = "Event time data as observed (L) versus to a data analyst (R)", fig.show = 'hold'}
library(knitr)
structureEventTime <- c("../figures/censoring_1.pdf", "../figures/censoring_2.pdf")
include_graphics(structureEventTime)
```


### Characteristics of event time data

- `Individuals' do not all enter the study at the same time. 
    - This is referred to as *staggered entry*.
- When the study ends, some individuals still haven't had the event.
- Other individuals drop out or are lost during the study.
     - The last time they were still "free" of the event is all that is known.

The last two features relate to *censoring* of the failure times. 

The first of the times until the study ends or the subject drops out is called a *censoring time*.

### A hypothetical example

Assume 10 subjects assigned to each of four treatments after cancer remission, followed  until death or end of study at 36 months.  

The times to death (months):

**Trt 1:**
Deaths at 2, 3, 7, 9, 15, 16  (additional 4 alive at end)

**Trt 2:** 
Deaths at 1, 1, 2, 4, 4, 6, 7, 9, 11 (add. 1 alive at end)

**Trt 3:**
Deaths at 1, 1, 2, 4, 4, 5  (add. 4 alive at end)

**Trt 4:** 
Deaths at 2, 3, 7, 9, 15, 22, 27, 28, 29 (add. 1 alive at end)

- For Treatment 1, what is the average time to death?
- Comparing Treatments 1 and 2, which appears better?
- Comparing Treatments 1 and 3, which appears better?
- Comparing Treatments 3 and 4, which appears better?

### Descriptive comparisons of ``average'' death time 

\begin{tabular}{lccc}
\hline
Treatment & \multicolumn{2}{c}{Among Deaths:} & Median adjusting \\
\cline{2-3} 
 Group & Mean & Median & for Censoring (KM) \\ \hline
1 & 8.67 & 8.00 & 15.5\\
2 & 5.00 & 4.00 & 5.0 \\
3 & 2.83 & 3.00 & 4.5 \\
4 & 15.78 & 15.00 & 18.5\\ \hline
\end{tabular}

### Comparisons of treatments (p-values)

Methods to do these calculations coming in this course.

\begin{tabular}{lccc}
\hline
Comparison & Log-Rank Test & Wilcoxon Test & Exponential Model \\ \hline
1 vs 2     & 0.045 & 0.048 & 0.014 \\
3 vs 4     & 0.62 & 0.67 & 0.56 \\
1 vs 3     & 0.63 & 0.37 & 0.73 \\
2 vs 4     & 0.15 & 0.06 & 0.09 \\
1 vs 4     & 0.37 & 0.68 & 0.34 \\
\hline
\end{tabular}


### More key features of survival data

- Survival times are often *right-skewed*, so the median is usually a better measure of center than the mean.

- The median can be estimated from data that include censored observations (not always possible with the mean).  

- Calculating summary statistics and comparing survival distributions must account for cases without events.

- Comparisons between survival distributions may yield different conclusions depending on assumptions.    


### Types of censoring

**Right-censoring** 

Let $T_i$ be the time to event and $U_i$ be the time to censoring for an individual $i$. Only the r.v. $X_i=\min(T_i,U_i)$ is observed,  due to

- loss to follow-up
- drop-out
- study termination (`administrative censoring')

Called *right*-censoring because the true unobserved event
is to the right of the censoring time (i.e., after the censoring time).

In addition to $X_i$, the *failure indicator* $\delta_i$ is observable:

\[\delta_i=\left\{\begin{array}{ccc}
1 & \mbox{if} & T_i\le U_i \\
0 & \mbox{if} & T_i>U_i \end{array} \right. \]

### Types of censoring \ldots

**Left-censoring**

The r.v.'s observed are $Y_i=\max(T_i,U_i)$ and the failure indicator $\epsilon_i$:
\[\epsilon_i=\left\{\begin{array}{ccc}
1 & \mbox{if} & U_i\le T_i \\
0 & \mbox{if} & U_i>T_i \end{array} \right. \]

Ex.: study of age at which African children learn a task (Miller). 

- Some already knew the task (left-censoring) 
- Some learned during study (exact)
- Some had not yet learned by end of study (right-censoring)

### Types of censoring\ldots

**Interval-censoring**

Observe $(L_i,R_i)$ where  $T_i\in (L_i,R_i)$

Examples:

- Time to prostate cancer, observe longitudinal PSA measurements

- Time to undetectable viral load in AIDS studies, based on measurements of viral load taken at each clinic visit

- Detect recurrence of colon cancer after surgery. Follow patients every 3 months after resection of primary tumor.

This short course is restricted to right-censoring.

### Independent versus informative censoring

Censoring is **independent** if $U_i$ is independent of $T_i$.

Examples:

- If $U_i$ is the planned end of the study (say, 2 years after the study opens), then it is usually independent of the event times.
    - What if there is a trend over calendar time in the survival times?

- If $U_i$ is the time that a patient drops out of the study because they have become much sicker and/or had to discontinue taking the study treatment, then $U_i$ and $T_i$ are probably not independent.

An individual censored at $U$ should be *representative* of all subjects who survive to $U$.

Censoring is considered **informative** if the distribution of $U_i$ contains information about the parameters characterizing the distribution of $T_i$.

# Survival Distributions

### Some mathematical definitions

There are several equivalent ways to characterize the probability distribution of a survival random variable.

- The density function $f(t)$
- The survivor function $S(t)$
- The hazard function $\lambda(t)$
- The cumulative hazard function $\Lambda(t)$

Some are special to survival analysis.

### Density function 

For a *discrete* random variable

Suppose that $T$ takes values in $a_1,a_2, \ldots, a_J$.
\begin{align*}
 f(t) &=  P(T = t) \\
 &=  \left\{ \begin{array}{ccc}
 f_j &\mbox{ if } & t=a_j, j=1,2,\ldots,J \\[1ex]
 0 &\mbox{ if } & t\ne a_j, j=1,2,\ldots,J
 \end{array}
 \right.
 \end{align*}

For a *continuous* random variable

\[  f(t) = \lim_{\Delta t \rightarrow 0}
 \frac{1}{\Delta t}  P(t \le T < t+\Delta t) \]


### Survivorship function $S(t)$

In other settings, the cumulative distribution function
\[F(t)=P(T\le t) = 1- S(t)\]
is of interest.  

In survival analysis, interest tends to focus on $S(t) = P(T > t)$, the *survival* or *survivorship* function.\footnote{Be careful:  some books use the definition $S(t)=P(T \geq t)$.} 

The survival function measures the probability an individual will experience the event beyond time *t*.

### Survivorship function $S(t)$ \ldots

For a *continuous* random variable:
\[   S(t) = \int_{t}^{\infty} f(u) du \]

For a *discrete* random variable:
\[ S(t) = \sum_{u > t} f(u) = \sum_{a_j > t} f(a_j) = \sum_{a_j > t} f_j \]

### Hazard function $\lambda(t)$

The hazard function measures the probability of death at time $t$, conditional on having survived until that time. 

  - This is sometimes called the *instantaneous failure rate*. 

For a *continuous* random variable $T$:
\begin{align*}
\lambda(t)  &= \lim_{\Delta t \rightarrow 0} \, \frac{1}{\Delta t}\,
P(t \le T <  t+\Delta t | T \ge t) \\[2ex]
&= \lim_{\Delta t \rightarrow 0} \, \frac{1}{\Delta t} \,
\frac{P([t \le T <  t+\Delta t] ~~ \bigcap~~ [T \ge t])}
{P(T \ge t)}\\[2ex]
&= \lim_{\Delta t \rightarrow 0} \, \frac{1}{\Delta t} \,
\frac{P(t \le T <  t+\Delta t)}{P(T \ge t)}\\[2ex]
&= \frac{f(t)}{S(t)}
\end{align*}

### Hazard function $\lambda(t)$\ldots

For a *discrete* random variable $T$:

\begin{align*}
  \lambda(a_j)\equiv \lambda_j   &= P(T = a_j | T \ge a_j)
   =  \frac{P(T=a_j)}{P(T\ge a_j)}\\[1ex]
  &= \frac{f(a_j)}{S(a_j -)}
  = \frac{f(t)}{\sum_{k:a_k\ge a_j}f(a_k)}
  \end{align*}

The form of the denominator for both continuous and discrete variables is the reason some books use $P(T \ge t)$ as the definition of the survivor function.


### Cumulative hazard function $\Lambda(t)$

For a *continuous* random variable *T*:
\begin{align*}
\Lambda(t)  &=\int_0^t \lambda(u) du
\end{align*}

For a *discrete* random variable *T*:
\begin{align*}
\Lambda(t) &= \sum_{k:a_k \leq t} \lambda_k
\end{align*}

### Relationship between $S(t)$ and $\lambda(t)$

For a continuous random variable,
\[
\lambda(t)=\frac{f(t)}{S(t)}.
\]

For a right-continuous survivor function $S(t)$,
\[
f(t) = -S'(t) \text{ or } S'(t) = -f(t).
\]

These relationships can be used to show another way to write $\lambda(t)$:
\[
\lambda(t) = \frac{f(t)}{S(t)} = - \frac{-f(t)}{S(t)} = - \left(\frac{1}{S(t)} \right) S'(t) = - \frac{d}{dt} [\log S(t)]
\]

\[
\lambda(t)= - \frac{d}{dt} [\log S(t)]
\]

### Relationship between $S(t)$ and $\Lambda(t)$

For a *continuous* random variable:

\begin{align*}
\Lambda(t) &= \int_0^t \lambda(u)du \\
&= \int_0^t -\frac{d}{du} \log S(u) du \\
&= -\log S(t)+\log S(0)\\
\end{align*}

Thus, $S(t)=e^{-\Lambda(t)}$.

### Relationship between $S(t)$ and $\Lambda(t)$\ldots

For a *discrete* random variable: 

Suppose that $a_j\leq t < a_{j+1}$.  Then
\begin{align*}
S(t)= P(T>t ) &= P(T > a_1, T > a_2, \ldots, T >  a_{j}) \\
&= P(T >  a_1) \times P(T >  a_2|T > a_1) \times \cdots \\
&\,\, \cdots \times P(T >  a_{j}|T > a_{j-1})\\
&= (1-\lambda_1) \times \cdots \times (1-\lambda_j)\\
&= \prod_{k:a_k \leq t}(1-\lambda_k)
\end{align*}

### Relationships: an overview

$$  f(t) \Delta t  \approx  P(t \le T < t+\Delta t) $$ 
$$ \lambda(t) \Delta t \approx  P(t \le T < t+\Delta t | T \ge t) $$
$$   S(t) = P(T>t) = \int_{t}^{\infty} f(u) du $$  
$$ f(t) =  - \frac{d}{dt} S(t) $$ 
$$ \lambda(t)=\frac{f(t)}{S(t)} $$ 
$$ \lambda(t)  =  - \frac{d}{dt}  [\log S(t)] $$ 

\[ S(t) = e^{-\Lambda(t)} \]
\[S(t) \approx 1 -\Lambda(t) \text{ while the cumulative hazard is small.}\]

### Some parametric survival distributions

The *exponential distribution* 

- Simplest distribution, only one unknown parameter
- Plays a role similar to that of the normal distribution in 
linear regression

\begin{align*}
f(t) &= \lambda e^{-\lambda t} \mbox{ for } t\ge 0 \\
S(t) &=  \int_t^\infty f(u) du  =  e^{-\lambda t} \\
\lambda(t) &=  \frac{f(t)}{S(t)}  = 
\lambda ~~~~~\mbox{ constant hazard} \\
\Lambda(t) &=  \int_0^t \lambda(u) \, du  = 
\int_0^t \lambda \, du  = \lambda t  
\end{align*}


### Some parametric survival distributions\ldots

The *Weibull distribution* generalizes the exponential, and has two parameters

-  $\lambda$:  the *scale* parameter    
-  $\gamma$: the *shape* parameter

\begin{align*}
S(t) &= e^{-\lambda t^\gamma} \\
f(t) &=  \frac{-d}{dt}S(t)=\gamma \lambda t^{\gamma-1}
e^{-\lambda t^\gamma}\\
\lambda(t) &= \gamma \lambda t^{\gamma-1}\\
\Lambda(t) &= \int_0^t\lambda(u) du = \lambda t^\gamma
\end{align*}

### The Weibull distribution\ldots

The Weibull distribution is convenient because of simple forms.  It includes several hazard shapes:

-  $\gamma=1 \rightarrow \mbox{ constant hazard}$ 

-  $0<\gamma<1 \rightarrow \mbox{ decreasing hazard}$ 

-  $\gamma > 1 \rightarrow \mbox{ increasing hazard}$

<!--- insert figures here --->

