---
title: "Significance Testing"
author: "Dave Harrington"
date: "May 14 - 18, 2018"
output: 
  beamer_presentation:
     includes:
       in_header: ../survival_header.tex
     fig_width: 3.25
     fig_height: 3.0
     fig_caption: true
     toc: true
     keep_tex: true
slide_level: 3
urlcolor: darkblue 
linkcolor: darkblue
citecolor: darkblue

--- 

# Introduction    

### Significance testing with event-time data

In the medical literature, survival analysis is frequently used to analyze clinical trials that may potentially change practice.    

- Significance testing is particularly important in this setting.

The figure on the next slide appeared at the beginning of Unit 1.

- Shows results of a randomized trial of ablation versus drug treatment for atrial fibrillation: 

    - Estimates of probability of survival or hospital admission by treatment group 

    - A $p$-value based on a *log-rank* test

This unit explores log-rank tests and other testing methods.



### Example: Time to death or hospitalization

![Figure from Marrouche, et al., *NEJM* 2018](../figures/atrial_fib_death_hosp.pdf){width=70%}


### Example: Clinical trial PBT01

![Figure from Stadtmauer, et al, NEJM 2000](../figures/pbt01_survival.pdf){width=70%}

The next slides reproduce this figure and $p$-value from patient-level data.

### Numerical summary

\scriptsize
```{r, echo=TRUE, eval=FALSE, fig.width=4.5, fig.height=3.3}
library(survival)
library(eventtimedata)
data("pbt01")

pbt01.survival <- survfit(Surv(survival, died) ~ treatment, 
                              data = pbt01)
pbt01.survival
```

### Numerical summary \ldots

\scriptsize
```{r, echo=FALSE, eval=TRUE, fig.width=4.5, fig.height=3.3}
library(survival)
library(eventtimedata)
data("pbt01")

pbt01.survival <- survfit(Surv(survival, died) ~ treatment, 
                              data = pbt01)
pbt01.survival
```

### Figure

\scriptsize
```{r, echo=TRUE, eval=FALSE, fig.width=4.5, fig.height=4.0}
library(survival)
library(eventtimedata)
data("pbt01")

pbt01.logrank.chisq = survdiff(Surv(survival, died) ~ treatment, 
                                data = pbt01)$chisq
pbt01.logrank.pvalue = pchisq(pbt01.logrank.chisq, 1,
                              lower.tail = FALSE)

plot(pbt01.survival, lty = 2:3, col = 3:4, mark.time = TRUE, 
     xlab = "Months after Randomization", 
     ylab = "Probability of Survival",
     axes = FALSE,
     cex = 0.6)
legend(40, 1.0, c("ABMT", "Control"), lty = 2:3, col = 3:4,
       cex = 0.6)
text(40, 0.6, "log-rank p-value = ", cex = 0.6)
text(55, .6, round(pbt01.logrank.pvalue, digits = 2), cex = 0.6)
text(70, 0.6, "(unstratified)", cex = 0.6)
axis(1)
axis(2)
```

----

\scriptsize
```{r, echo=FALSE, eval=TRUE, fig.width=4.5, fig.height=4.0}
library(survival)
library(eventtimedata)
data("pbt01")

pbt01.logrank.chisq = survdiff(Surv(survival, died) ~ treatment, 
                                data = pbt01)$chisq
pbt01.logrank.pvalue = pchisq(pbt01.logrank.chisq, 1,
                              lower.tail = FALSE)

plot(pbt01.survival, lty = 2:3, col = 3:4, mark.time = TRUE, 
     xlab = "Months after Randomization", 
     ylab = "Probability of Survival",
     axes = FALSE,
     cex = 0.6)
legend(40, 1.0, c("ABMT", "Control"), lty = 2:3, col = 3:4,
       cex = 0.6)
text(40, 0.6, "log-rank p-value = ", cex = 0.6)
text(55, .6, round(pbt01.logrank.pvalue, digits = 2), cex = 0.6)
text(70, 0.6, "(unstratified)", cex = 0.6)
axis(1)
axis(2)
```

\footnotesize

Note: The $p$-value is not identical to the one in the paper because the paper used a stratified test, stratifying on the cycle needed to induce complete response for the patient. Stratified tests coming later.

### Parametric vs non-parametric approaches

In the medical literature, survival analysis is often used in the study of treatments for chronic diseases such as cancer, diabetes, or cardiovascular disease.

In most studies, a proportion of participants have not had an event by the time the study is analyzed.    

- Thus, the right tail of the survival distribution is not observable.

Parametric approaches can be useful in some settings, but they assume a model for the entire curve, and extrapolate tail behavior.

Non-parametric methods make no assumptions about tail behavior and are less sensitive to outliers.

This section emphasizes *non-parametric methods*.

### Two-sample non-parametric tests for comparing survival distributions

Comparing two distributions at a single time point

The log-rank test

Generalized Wilcoxon tests

The Fleming-Harrington family

\vspace{1cm}

\small

References:

\begin{tabular}{ll}
Hosmer \& Lemeshow &  Section 2.4\\
Collett & Section 2.5\\
Klein \& Moeschberger~~~~ & Section 7.3\\
\end{tabular}

# Comparing two distributions at a single time point      

### Using  $\widehat{S_1}(t)$ and $\widehat{S_2}(t)$ 

Sometimes a specific time point, $t^{\star},$ is of special interest.     

- e.g., 5-year disease-free survival in cancer

Simple method:       

- Use the independence and approximate normality of $\widehat S_k(t^\star);  k \in\{ 0,1 \}$.   

- Examine a confidence interval for the difference in estimated survival curves at $t^{\star}$.

- Reject $H_0: S_1(t) = S_2(t)$ in favor of a two-sided alternative if the interval does not include 0.

### Confidence interval for the difference of two survival curves

The 95\% confidence interval is 
\[
\left[ \left ( \widehat S_1(t^\star) - \widehat S_0(t^\star) \right )
\pm 1.96 \times \sqrt{ V_1(t^\star) + V_0 (t^\star) }\right],
\]
where $V_k(t^\star)$ is the estimated variance of $\widehat S_k(t^\star).$

This method is rarely used because

- it is not clear what $t^\star$ should be

- there is potential for abuse when applied *post-hoc*

### Example

Use numerical estimates of the survival curves to find a 95\% confidence interval for the difference in survival curves at time point $t^\star = 10$ weeks for the Cox and Oakes leukemia data.

The following slides show the estimates repeated from Unit 2 (Estimation).

### KM numerical estimates, group == 0

\scriptsize

```{r echo = TRUE, eval=TRUE}
library(survival)
library(eventtimedata)
leukemia.group.0 = 
  subset.data.frame(cox.oakes.leukemia, group == 0)
km.group.0 = survfit(Surv(time, relapse) ~ 1, 
                     data = leukemia.group.0)
summary(km.group.0)
```

### KM numerical estimates, group == 1

\scriptsize

```{r echo = TRUE, eval=TRUE}
leukemia.group.1 = 
  subset.data.frame(cox.oakes.leukemia, group == 1)
km.group.1 = survfit(Surv(time, relapse) ~ 1, 
                     data = leukemia.group.1)
summary(km.group.1)
```
\normalsize

### Calculations \ldots

Lab exercise!

# The log-rank test   

### Mantel-Haenszel log-rank test

The log-rank test is the most widely used non-parametric test.

Begin with a  $2 \times 2$ table classifying those with and without the event of interest in a two group setting:
\begin{center}
\begin{tabular}{cccc}
\hline \hline
& \multicolumn{2}{c}{Event} & \\ \cline{2-3}
\multicolumn{1}{c}{Group } & ~~~~Yes~~~~ & ~~~~No~~~~ & ~~~Total~~~\\ \hline
0 & $d_0$ & $n_0 - d_0$ & $n_0$  \\
1 & $d_1$ & $n_1 - d_1$ & $n_1$ \\
\hline
Total &  $d  $ & $n   - d  $ & $n  $  \\ \hline \hline
\end{tabular}
\end{center}

The table shows the observed numbers with and without events
in each group, and the margin totals.  

### Mantel-Haenszel approach to a $2\times2$ table

Define $D_0$ as the random variable representing the number with an event in Group 0.

If the margins of this table $(d, n-d, n_0, n_1)$ are considered fixed, then $D_0$ follows a hypergeometric distribution,  depending on one parameter (the population odds ratio, $\psi$).

Under $H_0$, the null hypothesis of no association between the event and group:
\[E(D_0) = \frac{n_0 \, d}{n} = n_0 \left(\frac{d}{n}\right) \]

\[\text{Var}(D_0) = \frac{n_0 \, n_1 \, d(n-d)}{n^2 (n-1)} \]

### Mantel-Haenszel approach to a $2\times2$ table\ldots

Thus, the Mantel-Haenszel statistic is
\[\chi^2_{\tiny MH } = \frac{\left[d_0-n_0 \, d/n\right]^2}{\frac{n_0
\, n_1 \, d(n-d)} {n^2 (n-1)}} \sim \chi^2_1
\]

$\chi^2_{\tiny MH}$ is approximately equivalent to the Pearson $\chi^2$ test for equality of the two groups given by:
\[\chi^2_p = \sum \frac{(o-e)^2}{e}, \]
where $o$ represents observed values and $e$ the expected values.

### Example: Toxicity in a clinical trial with two treatments

\begin{center}
\begin{tabular}{cccc}
\hline \hline
& \multicolumn{2}{c}{Toxicity}   \\ \cline{2-3}
\multicolumn{1}{c}{Group } & ~~~Yes~~~ & ~~~No~~~ & ~~~Total~~~ \\ \hline
0 &  ~8    & 42  &  ~50       \\
1 &  ~2   &  48 &   ~50        \\
\hline
Total & 10 & 90 &  100        \\ \hline \hline
\end{tabular}
\end{center}


\begin{align*}
\chi_p^2 &= 4.00 ~~~~ (p=0.046)\\[3ex]
\chi_{\tiny MH}^2 &= 3.96 ~~~~ (p=0.047)
\end{align*}


### Pearson $\chi^2$ vs MH

*Note:* the Pearson $\chi^2$ test applies to the case where the row margins are fixed but not the column margins, as a test of equivalence between the proportions with events in the two groups.  

In this case, the variance is slightly different: 
\[
\text{Var}(D_0) = \frac{n_0 \, n_1 \, d(n-d)}{n^3}
\]

### For the case of $K$ tables

Now suppose there are $K$ $2 \times 2$ tables, all independent. 

The goal is  to test for a common group effect $H_0: \psi_j=\psi=1$ 
versus $H_A: \psi \neq 1$. 

The *Cochran-Mantel-Haenszel test* for a common odds ratio not equal to 1 can be written as:

\[  
\chi^2_{CMH} =   \frac{\left[ {\sum_{j=1}^K (d_{0j} - n_{0j} \times d_j/n_j)}\right]^2}
        {\sum_{j=1}^K n_{1j} n_{0j} d_j (n_j-d_j)/[n_j^2(n_j-1)]}  \]


This statistic is distributed approximately as $\chi^2_1$.

### $K$ tables \ldots

The subscript $j$ refers to the $j$-th table:

\begin{center}
\begin{tabular}{cccc}
\hline \hline
& \multicolumn{2}{c}{Event} & \\ \cline{2-3}
\multicolumn{1}{c}{Group } & ~~~Yes~~~ & ~~~No~~~ & ~~~Total~~~\\ \hline
0 & $d_{0j}$ & $n_{0j} - d_{0j}$ & $n_{0j}$  \\
1 & $d_{1j}$ & $n_{1j} - d_{1j}$ & $n_{1j}$ \\
\hline
Total &  $d_j  $ & $n_j - d_j  $ & $n_j  $  \\ \hline \hline
\end{tabular}
\end{center}

### Log-rank Test: Applying CMH to survival data

For the two-group *log-rank* test:    

- Construct a $2 \times 2$ table at each distinct failure time.

- Compare the failure rates between the two groups, conditional on the number at risk in the groups.

- Combine the results from each table using the Cochran-Mantel-Haenszel test.

### Formal notation for the log-rank test

Let $t_1, \ldots ,t_K$ represent the $K$ ordered, distinct failure times.

The table at the $j$-th failure time, is
\begin{center}
\begin{tabular}{cccc}
\hline \hline
& \multicolumn{2}{c}{Die/Fail} & \\ \cline{2-3}
\multicolumn{1}{c}{Group } & ~~~Yes~~~ & ~~~No~~~ & ~~Total~~\\ \hline
0 & $d_{0j}$ & $r_{0j} - d_{0j}$ & $r_{0j}$ \\[2ex]
1 & $d_{1j}$ & $r_{1j} - d_{1j}$ & $r_{1j}$ \\[2ex]
\hline
Total &  $d_j$ & $r_j - d_j$ & $r_j$  \\ \hline \hline
\end{tabular}
\end{center}
where    

- $d_{0j}$ and $d_{1j}$ are the number of failures in group 0 and 1, respectively, at the $j$-th failure time   

-  $r_{0j}$ and $r_{1j}$ are the number at risk in groups 0 and 1, at the $j$-th failure time

### The log-rank test statistic formula

\begin{align*}
\chi^2_{\text{log-rank}} &=
\frac{\left[{\sum_{j=1}^K (d_{0j} - r_{0j} \times d_j/r_j)}\right]^2}
{\sum_{j=1}^K  \frac{r_{1j} r_{0j} d_j (r_j-d_j)}{[r_j^2(r_j-1)]}}
\end{align*}

If the tables are all independent, then this statistic will have an approximate $\chi^2$ distribution with 1 df.


### Notes about log-rank tests

The log-rank statistic depends on ranks of event times only, that is, on the order in which events and censorings occur.

If there are no tied failure times between the two groups, then $d_j=1$ and the log-rank statistic simplifies to
\[\chi^2_{\text{log-rank}}  =  \frac{[\sum_{j=1}^K (d_{0j} - \frac{r_{0j}}{r_j})]^2}
        {\sum_{j=1}^K r_{1j}r_{0j}/r_j^2}  \]

The numerator can be interpreted as $[\sum (o-e)]^2$, where    

- $o$ is the observed number of deaths in a group, and $e$ is the expected number, given the risk sets.

- The expected number equals the number of failures times the proportion at risk in the group.

- It does not matter which group is used for the sum.

### Notes about the log-rank\ldots

The ($o-e$) terms in the numerator  can be written as
\[   \frac{r_{0j}r_{1j}}{r_j}(\widehat\lambda_{1j} - \widehat\lambda_{0j} ) \]

Solution as a lab problem!

### Assumptions behind log-rank test

Censoring is independent.

- This assumption is made in nearly all survival methods.

The contributions to the statistic made by the $2 \times 2$ tables can be treated as independent.

- Proven true in the 1980's

The log-rank test is most powerful when hazards have a constant ratio over time.

- This is termed the *proportional hazards* assumption.

- It is not required for validity under the null hypothesis.




### Efficiency of the log-rank test

The CMH test for a series of tables stratified by a potential confounder is most powerful when \ldots  

- The tables have a constant odds ratio.

Analogously, the log-rank test is most powerful when \ldots

- The hazard ratios are constant across $t$ time intervals.  

- This corresponds to *proportional hazards*.

### The recidivism data

```{r, eval = TRUE, echo=FALSE, fig.width=4.5, fig.height=3.8}
library(survival)
library(eventtimedata)
data("rossi")
rossi.recidivism.km <- survfit(Surv(week, arrest) ~ fin, data = rossi)
plot(rossi.recidivism.km, lty = 2:3, col = 3:4, mark.time = TRUE, 
     xlab = "Weeks", 
     ylab = "Probability of Re-arrest", 
     axes = FALSE, 
     conf.times = c(10,30,50),
     main = "KM of Recidivism Probability, with Conf. Int.",
     cex = 0.6, cex.main = 0.8)
axis(1)
axis(2)
legend(10, .5, c("Financial Support", "No Financial Support"), lty = 2:3, col = 3:4,
       cex = 0.6)
```
     
### The recidivism data \ldots

\scriptsize

```{r, echo=TRUE, eval=TRUE, fig.width=4.5, fig.height=3.3}
library(survival)
library(eventtimedata)
data("rossi")
survfit(Surv(week, arrest) ~ fin, 
                        data = rossi)
survdiff(Surv(week, arrest) ~ fin, 
                        data = rossi)
```

### What does non-proportional hazards look like?

The next slides show figures presented at a February 5, 2018 meeting on non-proportional hazards. 

All use data from published papers.

The workshop (sponsored by Duke University Margolis Center)

- reviewed instances where non-proportional hazards occurred in studies designed for drug approval

- discussed strategies for modifying usual methods of analysis

### Randomized trial in prostate cancer

![Data from Hussain, et al., *Euro J Cancer*, 2015](../figures/hussain.pdf){width=80%}

See [Hussain, et al.](../../clinical_papers/hussain.pdf)


### RCT in acute leukemia

![KM survival curves from Kantarjian, et al., *NEJM*, 2016](../figures/kantarjian.pdf){width=80%}

See [Kantarjian, et al.](../../kantarjian.pdf)


### RCT in acute leukemia \ldots

![Cumulative hazards from Kantarjian, et al., *NEJM*, 2016](../figures/cum_haz_ALL.pdf){width=80%}

### Another example, but with data

\scriptsize
```{r}
library(devtools)
install_github("keaven/nphsim")
library(survival)
library(nphsim)
data(Ex6crossing)
survfit(Surv(month,evntd) ~ trt, data = Ex6crossing)
survdiff(Surv(month,evntd) ~ trt, data = Ex6crossing)
```

### The Kaplan-Meier estimate

```{r, eval=TRUE, echo=FALSE, fig.width=4.5, fig.height=3.3}

library(survival)
library(nphsim)
data(Ex6crossing)

km = survfit(Surv(month,evntd) ~ trt, data = Ex6crossing)
plot(km, lty = 2:3, col = 3:4, mark.time = TRUE,
     xlab = "Months",
     ylab = "Survival Probability",
     axes = FALSE)
axis(1)
axis(2)
```

### The cumulative hazard

```{r, eval = TRUE, echo=FALSE, fig.width=4.5, fig.height=3.8}
library(survival)
library(nphsim)
data(Ex6crossing)

km <- survfit(Surv(month, evntd) ~ trt, data = Ex6crossing)
plot(km, lty = 2:3, col = 3:4, mark.time = TRUE, 
     fun = "cumhaz",
     xlab = "Months", 
     ylab = "Cumulative Hazard", 
     axes = FALSE, 
     cex = 0.6, cex.main = 0.8)
axis(1)
axis(2)
```

# Weighted log-rank tests   

### The Tarone-Ware class of tests

\small
This general class of tests is like the log-rank test, but adds weights ${w_j}$.  

Many specific test statistics are included as special cases.
\[ \chi^2_{TW} = \frac{[\sum_{j=1}^K {w_j} (d_{1j} - r_{1j} \times d_j/r_j)]^2}
        {\sum_{l=1}^K \frac{w_j^2 r_{1j}r_{0j}d_j(r_j-d_j)}{r_j^2(r_j-1)}}  \]

\begin{center}
\begin{tabular}{ll}
\hline \hline
Test statistic &  Weight ${w_j}$ \\ \hline
Log-rank              &  $w_j = 1$\\[2ex]
Gehan's Wilcoxon     &  $w_j = r_j$\\[2ex]
Peto/Prentice Wilcoxon       &  $w_j = n \widehat{S}(t_j)$\\[2ex]
Fleming-Harrington~~~~   &  $w_j=[\widehat{S}(t_j)]^\rho ~ [1-\widehat{S}(t_j)]^\gamma$\\[2ex]
Tarone-Ware          &  $w_j=\sqrt{r_j}$\\
\hline \hline
\end{tabular}
\end{center}

$r_j$ is the number of subjects at risk at the $j^{th}$ event time.

### Some background

The generalized Wilcoxon tests precede the Tarone-Ware or Fleming-Harrington class of tests.

 - The Gehan-Wilcoxon was derived using a generalization of the $U$ statistic approach to the Mann-Whitney-Wilcoxon.

 - The Peto/Prentice Wilcoxon was derived using a generalization of linear rank statistics.

### More details on the Fleming-Harrington test

The parameters $\rho$ and $\gamma$ can be any non-negative numbers:

- If $\rho = \gamma = 0$, $w_j=1$ and the test is the usual log-rank test.    
  
-  If $\rho=1$ and $\gamma=0$, the test is similar to the Peto-Prentice.\footnote{This is the default ``Fleming'' test in SAS PROC LIFETEST.}

- If $\rho=0$ and $\gamma=1$, what happens to $w_j$ over follow-up time?

- If $\rho = \gamma = 1$, the weight $w_j$ reaches a maximum at the median, and is smaller for both large and small $t_j$.

The \texttt{survdiff()} function in \textsf{R} sets $\gamma=0$ and allows the user to set $\rho$.

### Earlier numerical example, $\rho = 1$

\small

This is a generalized Wilcoxon test

```{r}
#library(devtools)
#install_github("keaven/nphsim")
library(survival)
library(nphsim)
data(Ex6crossing)
survdiff(Surv(month,evntd) ~ trt, rho = 1, data = Ex6crossing)
```

### Earlier numerical example, $\rho = 2$

\small

```{r}
#library(devtools)
#install_github("keaven/nphsim")
library(survival)
library(nphsim)
data(Ex6crossing)
survdiff(Surv(month,evntd) ~ trt, rho = 2, data = Ex6crossing)
```

### Be careful with weighted LR tests

The weighted LR tests are often presented as emphasizing differences between two hazard functions.   

For the F-H weights, $(\widehat{S}(t))^\rho(1 - \widehat{S}(t))^\gamma$,

- $\rho > 0,\,\, \gamma = 0$: weights early differences ($\rho = 1$ gen Wilcoxon)    

- $\rho = 0,\,\, \gamma > 0$: weights late differences

- $\rho > 0,\,\, \gamma > 0$: weights differences near median

- $\rho = 0,\,\, \gamma = 0$: weights differences equally over time (log-rank)

Choosing the test post-hoc, based on observed data, leads to potentially increased Type I error.

The [February 2018 Duke-Margolis workshop](https://healthpolicy.duke.edu/events/public-workshop-oncology-clinical-trials-presence-non-proportional-hazards) discussed ways to specify these tests in design and sample size calculations.

- Full workshop materials available at the link.


# Tests for more than two groups

### Introduction

Suppose data come from $P$ different groups. The data from group $p$ ($p=1,\ldots,P$) are:
\[ (X_{p1},\delta_{p1}) \dots
(X_{p n_p},\delta_{p n_p}) \] 

Tests are based on a $P \times 2$ table at each distinct $K$ failure time.

- Compare the event rates between the $P$ groups, conditional on the number at
risk, combining the tables using the CMH approach

- Final test statistic has $\chi^2$ distribution with $P - 1$ degrees of freedom

### Example: prognosis in lymphoma

The data \texttt{lymphoma.prognosis} in the package \texttt{eventtimedata} was used as the training sample in the International Prognostic Index published by [Shipp, et al.](../../clinical_papers/shipp_int_prog_index_nejm.pdf) in 1993. 

The data record survival time and censoring for 1,385 patients with non-Hodgkin's lymphoma treated at sites in the US, Canada, and Europe.

In this analysis, we look at the association of disease stage and survival.  See the package documentation for the variable definitions.

### The code for the 4-group test

\footnotesize

```{r, eval=FALSE, echo=TRUE}
library(survival)
library(eventtimedata)
data(lymphoma.prognosis)

stage.factor = as.factor(lymphoma.prognosis$STAGE)
died = lymphoma.prognosis$SURVIVAL - 1
died[died == 2] = 0  #recoding those lost to follow-up as censored
survival.time = lymphoma.prognosis$SURVTIME
lymphoma.survival <- survfit(Surv(survival.time, died) ~ 
                               stage.factor)
lymphoma.survival
survdiff(Surv(survival.time, died) ~ stage.factor)
```

### The output for the 4-group test

\footnotesize
```{r, eval=TRUE, echo=FALSE}
library(survival)
library(eventtimedata)
data(lymphoma.prognosis)

stage.factor = as.factor(lymphoma.prognosis$STAGE)
died = lymphoma.prognosis$SURVIVAL - 1
died[died == 2] = 0  #recoding those lost to follow-up as censored
survival.time = lymphoma.prognosis$SURVTIME
lymphoma.survival <- survfit(Surv(survival.time, died) 
                             ~ stage.factor)
lymphoma.survival
survdiff(Surv(survival.time, died) ~ stage.factor)
```
### The survival plot


```{r, eval=TRUE, echo=FALSE, fig.width=4.5, fig.height=3.8}
library(survival)
library(eventtimedata)
data(lymphoma.prognosis)

stage.factor = as.factor(lymphoma.prognosis$STAGE)
died = lymphoma.prognosis$SURVIVAL - 1
died[died == 2] = 0  #recoding those lost to follow-up as censored
survival.time = lymphoma.prognosis$SURVTIME
lymphoma.survival <- survfit(Surv(survival.time, died) 
                             ~ stage.factor)

plot(lymphoma.survival, col = 1:4, mark.time = TRUE, 
     xlab = "Years", 
     ylab = "Survival Probability", 
     axes = FALSE, 
     main = "Survival by Stage",
     cex = 0.6, cex.main = 0.8)
axis(1)
axis(2)
legend(2, .4, c("Stage 1", "Stage 2", "Stage 3", "Stage 4"), lty = 1,
       col = 1:4, cex = 0.6)
```

# The stratified log-rank test

### Example: length of stay in a nursing home

The National Center for Health Services Research studied 36 for-profit nursing homes to assess 

-   effects of different financial incentives on length of stay

``Treated" nursing homes received 

- Higher daily reimbursements for US Medicaid (financially needy) patients

- Bonuses for improving a patient's health and sending them home

Study included 1601 patients admitted between May 1, 1981 and
April 30, 1982.\footnote{Data are in \texttt{nursing.home} in the \texttt{eventtimedata} package.}

### Differences in length of stay by treatment

\footnotesize

```{r, eval=TRUE, echo=TRUE}
library(survival)
library(eventtimedata)
data(nursing.home)
survdiff(Surv(stay, cens) ~ rx, data = nursing.home)
```
### A stratified analysis

Length of stay may also be associated with gender. 

- Women tend to be healthier in the US.

A stratified test allows one to test for treatment differences, adjusting for gender (without using a modeling approach).

 - assumes the shape of the hazard may vary between men and women, but that the effect of the incentive would be the same

 - easy to do in almost any software

### Differences in length of stay by treatment, stratified by gender

\footnotesize

```{r, eval=TRUE, echo=TRUE}
library(survival)
library(eventtimedata)
data(nursing.home)
survdiff(Surv(stay, cens) ~ rx + strata(gender), 
         data = nursing.home)
```

### The stratified test for the PBT01 data

Log-rank test stratified on \texttt{cycle.of.resp}.

This is the $p$-value in the Stadtmauer paper.

  - Unstratified $p$-value (shown in earlier slides) is 0.34.

\scriptsize
```{r, echo=TRUE, eval=TRUE}
library(survival)
library(eventtimedata)
data("pbt01")

survdiff(Surv(survival,died) ~ treatment + strata(cycle.of.resp),
              data = pbt01)
```

#  Derivations

### The $P$-group log-rank statistic

Let $t_1, \ldots, t_K$ represent the $K$ ordered, distinct failure times in the pooled sample.

At the $j$-th failure time, the following table summarizes the data,
\begin{center}
\begin{tabular}{cccc}
\hline \hline
& \multicolumn{2}{c}{Fail} & \\ \cline{2-3}
\multicolumn{1}{c}{Group } & ~~~Yes~~~ & ~~~No~~~ & ~~~Total~~~\\ \hline
1 &  $d_{1j}$  & $r_{1j} - d_{1j}$ & $r_{1j}$ \\[2ex]
. &    .       &     .             &       .    \\[2ex]
P & $d_{Pj}$   & $r_{Pj} - d_{Pj}$ & $r_{Pj}$  \\ \hline
Total &  $d_j  $ & $r_j - d_j  $ & $r_j  $  \\ \hline \hline
\end{tabular}
\end{center}
where $d_{pj}$ is  the number of deaths in group $p$ at the $j$-th failure time, and $r_{pj}$ is the number at risk at that time.

The tables are then combined using the CMH approach.


### Details of the calculation

\small
For one table at a particular failure time, the test statistic would be constructed from the $P \times 1$ vector of (observed - expected) values.

- Each group contributes one component of the sum.

\footnotesize

Let ${\bf O}_j = (d_1, \ldots, d_{(P-1)j})^T$ be a vector of the observed number of failures in groups 1 to $(P - 1)$ at the $j$-th death time. Given the risk sets $r_{1j}, \ldots, r_{Pj}$, and the fact that there are $d_j$ deaths, ${\bf O}_j$ has mean 
\[{\bf E}_j = \left(\frac{d_j r_{1j}}{r_j}, \ldots, \frac{d_j r_{(P-1)j}}{r_j} \right)^T \]

and variance-covariance matrix 

 \[   {\bf V}_j = \left( \begin{array}{cccc} 
                          v_{11j} & v_{12j} & ... & v_{1(P-1)j} \\
                                  & v_{22j} & ... & v_{2(P-1)j} \\
                              ... &         & ... & ... \\
                                  &         &     & v_{(P-1)(P-1)j} \\
                    \end{array} \right)  \]

### Details of the calculation\ldots

-  The $\ell$-th diagonal element is:
\[     v_{\ell\ell j} = r_{\ell j}(r_j-r_{\ell j}) 
d_j(r_j-d_j)/[r_j^2(r_j-1)]  \]

-  The $\ell m$-th off-diagonal element is:
\[     v_{\ell m j} = r_{\ell j}r_{mj}d_j(r_j-d_j)/[r_j^2(r_j-1)]  \]


### Details of the calculation \ldots

The resulting $\chi^2$ test for a single $P \times 1$ table has $(P-1)$ degrees of freedom and is constructed as follows:

\[   ( {\bf O}_j - {\bf E}_j)^T \; {\bf V}^{-1}_j \; ( {\bf O}_j - {\bf E}_j)
\]

To generalize to $K$ tables (i.e., $K$ failure times), combine as in the log-rank:

  - Let ${\bf O}_j$, ${\bf E}_j$ and ${\bf V}_j$ with the sums over the
$K$ distinct failure times. 

  - That is, let  ${\bf O} = \sum_{j=1}^{k}  {\bf O}_j$, ${\bf E} = \sum_{j=1}^{k} {\bf E}_j$, and ${\bf V} = \sum_{j=1}^{k}  {\bf V}_j$.  

The test statistic is:
\[
 ( {\bf O} - {\bf E})^T \; {\bf V}^{-1} \; ( {\bf O} - {\bf E}),
\]
and has a $\chi^2$ distribution with $P - 1$ degrees of freedom.

### The stratified log-rank test

Used when assessing the association between survival and a factor $X$ that has two different levels.  

- Want to stratify by a second factor, that has $S$ different levels.  

First, divide the data into $S$ separate groups.  

Within group $s$ ($s=1,...,S$), 

- Construct the usual log-rank to assess the association between survival and the variable $X$.  

- Let  $t_{1s}, \ldots ,t_{K_s s}$ represent the $K_s$ ordered, distinct death
times \underline{in the $s$-th group}.

### The stratified log-rank test \ldots

At the $j$-th death time in group $s$:  

\begin{center}
\begin{tabular}{cccc}
\hline \hline 
& \multicolumn{2}{c}{Die/Fail} & \\ \cline{2-3}
\multicolumn{1}{c}{X} & ~~~Yes~~~ & ~~~No~~~ & ~~~Total~~~\\ \hline
1 & $d_{s1j}$ & $r_{s1j} - d_{s1j}$ & $r_{s1j}$ \\[2ex]
2 & $d_{s2j}$ & $r_{s2j} - d_{s2j}$ & $r_{s2j}$  \\
\hline
Total &  $d_{sj}  $ & $r_{sj}   - d_{sj}  $ & $r_{sj}  $  \\ 
\hline \hline
\end{tabular}
\end{center}

### The stratified log-rank test \ldots

Let 

- $O_s$ be the sum of the ``o''s obtained by applying the log-rank calculations in the usual way to the data from group $s$.

- $E_s$ be the sum of the ``e''s, 

- $V_s$ be the sum of the ``v''s.  

The  *stratified logrank* test statistic is

\[  Z = \frac{\sum_{s=1}^{S} (O_s - E_s)}{\sqrt{\sum_{s=1}^{S} (V_s)}} \] 

The test can easily be extended to weighted log-rank tests and to more than two levels of the factor $X$.

<!---

Stadtmauer EA, O'Neill A, Goldstein PG, Crilley P, Mangan KF, Ingle JN, Lazarus HM, Erban J, Sickles C, Glick JH: Phase III Randomized trial of high-dose chemotherapy (HDC) and stem cell support (SCT) shows no difference in overall survival or severe toxicity compared to maintenance chemotherapy with cyclophosphomide, methotrexate and 5-fluorouracil (CMF) for women with metastatic breast cancer who are responding to conventional induction chemotherapy: The Philadelphia Intergroup Study (PBT-01), ECOG, U. of Pennsylvania. Proc Am Soc Clin Oncol 18: 1a, 1999 (abstr 1)

[HTML] Conventional-dose chemotherapy compared with high-dose chemotherapy plus autologous hematopoietic stem-cell transplantation for metastatic breast …
EA Stadtmauer, A O'Neill, LJ Goldstein… - … England Journal of …, 2000 - Mass Medical Soc
Background We conducted a randomized trial in which we compared high-dose 

> library("sas7bdat", lib.loc="/Library/Frameworks/R.framework/Versions/3.4/Resources/library")
> abmt <- read.sas7bdat("pbt01.sas7bdat")
> 

--->
